Extending K-Means Clustering with Ptolemy’s Inequality \thanks{This work is partly funded by DFG grant No. 454630593: EPIX – Efficient Ptolemaic Indexing.}

### 2024-11-14 content/3_preliminaries.tex
overleaf version as submitted

In this section, we first define the k-means clustering problem and the Lloyd algorithm as a solution to this problem in Section \ref{subsec: kmeans}.
We then continue with introducing means of acceleration via lower and upper bounding in Section \ref{sub:acc}.

@@

In order to discuss the \kmeans, it is sensible to define the concept of a \emph{clustering algorithm} formally.
\begin{definition}[clustering algorithm]
	Given a target number of clusters $k$,
	let $F$ be a function that assigns each element from an input data set $D\subset \mathbb{R}^n$ to a
	cluster index $i \in \{1, 2, \ldots, k\}$
	$$ F:D \to \{1, 2, \ldots, k\} \;.$$
\end{definition}
Notably, we restrict our analysis to clustering problems that use the $n$-dimensional Euclidean space as the domain of the input data set.
Likewise, we assume that the similarity of the data points to be clustered is completely defined by the Euclidean distance $d(x,y)= || x-y ||_2 $.
The rationale behind this assumption will be made clear shortly in \autoref{sub:acc}.

The heart of the \kmeans is the assignment of clusters so that the variance in similarity between all points in the same cluster should be reasonably small.
More formally:
\begin{definition}[\kmeans]
	Let $F$ be a clustering algorithm
	%$x \in D$ be an element from the input data set of a clustering algorithm $F$
	and let $C_i$ be the set of all elements assigned to cluster $i$:
	$$ C_i = \{x \in D \mid F(x) = i\} \,.$$
	Furthermore, let $\epsilon$ be the weighted sum of inter-cluster variances
	$$ \epsilon = \sum_{i=1}^k |C_i| \operatorname{Var}[C_i] = \sum_{i=1}^k \sum_{x_j \in C_i}  d^2(x_j, c_i)\,, $$
	where $c_i = \operatorname{E}[C_i]$ is center of cluster $C_i$.

	Then, the map $F$ is called a solution to the \kmeans
	if $\epsilon$ cannot be made smaller by changing the assignment of any one element to a different cluster.
\end{definition}
Note that with this definition, the common $k$-means algorithms provides solution to the \kmeans,
as we do not require that $F$ finds the smallest possible $\epsilon$,
just a local minimum of $\epsilon$.
This is consistent with the colloquial notion that solutions to the \kmeans ``can be found'',
even when an optimal solution (an NP-hard problem) is not practically obtainable \cite{han2012data}.

The most prevalent algorithm employed to identify valid solutions to the \kmeans is the Lloyd algorithm.
For convenience, it is reproduced in Algorithm~\ref{alg:lloyd};
given its widespread use, we will only cover its most significant characteristics as they relate to our analysis.

@@

For each iteration of Lloyds algorithm, the assignment step (\autoref{algstep:assign}) needs to identify the closest cluster center for each element from the input dataset $D$.
To do so, Lloyd's algorithm calculates the distances between each cluster and each element from $D$, requiring $k\cdot|D|$ total distance evaluations per iteration.
These frequent distance evaluations can make up a significant share of the algorithms computational costs, especially for high-dimensional datasets.

@@

As we have seen, Lloyd's algorithm requires numerous evaluations of the distance function.
However, the exact distances do not actually need to be calculated explicitly for every element, as it is sufficient to identify which cluster center is the closest to a given element.
This is often possible from geometric considerations alone, when lower and upper distance bounds are available.
Two properties of the Euclidean space are of particular use here:
The triangle inequality
\begin{align}
	\label{eq:tri}
	d(x,y) \leq d(x,z) + d(z,y)
\end{align}
and Ptolemy's inequality
\begin{align}
	\label{eq:pto}
	d(x, y)\cdot d(v, u) \leq d(x, v) \cdot d(y,v) + d(x, u) \cdot d(y, v)
\end{align}
can provide lower and upper bound on the distance between two points given two (respectively five) other distances.

Numerous algorithms have been developed that exploit the former inequality to avoid explicit distance evaluations. However, to the best of our knowledge, the latter has not been employed for the purposes of accelerating k-means clustering yet.

Therefore, we modify Elkan's algorithm, an existing method that already employs the triangle inequality, to also make use of the Ptolemaic inequality.
We provide a short summary of Elkan's algorithm \cite{DBLP:conf/icml/Elkan03} here and describe our extension in \autoref{sec:contrib}.

Among Elkan's algorithm maintains two kinds of distance bounds each iteration of the algorithm:
\begin{enumerate}[label=\roman*]
	\item $u(x) \geq d(x, c_i) \mid i = F(x)$,
	      upper bounds between data points and their currently assigned centers.
	\item $l(x, c_i) \leq d(x,c_i) \forall i \neq F(x)$,
	      lower bounds between points and all other centers.
\end{enumerate}
Intuitively, when the upper bound is larger than all lower bounds $u(x)\leq l(x,c_i) \forall i\neq F(X)$, the cluster assignment of a data point has not changed.
Explicit distances to other clusters are only required when this inequality does not hold, i.e. when the bounds are not tight enough or when the cluster assignment of the data point has indeed changed.
It is immediately evident that the quality of the bounds has a significant impact on the achievable improvements in execution speed:
The larger (smaller) the lower (upper) bound is, the more likely an explicit distance evaluation can be avoided.

While Elkan's algorithm also employs additional techniques to reduce the number of distance calculations, we want to focus on the calculation of the bounds here;
a complete description of the algorithm is given in Algorithm~\ref{alg:elkan}, as reproduced from \cite{elkan}.

The lower and upper bounds are calculated at the end of each iteration of Elkan's algorithm and are given by
\begin{align}
	\label{eq:elkan_lower}
	l(x_i, c_j) & = \max \{ l(x_i, c_j) - d(c'_j, c_j), 0 \} \\
	\label{eq:elkan_upper}
	u(x_i)      & = u(x_i) + d(c'(x_i), c(x_i)) \,,
\end{align}
where the prime symbol ($c'_j$) indicates the location of the new cluster centers, while unprimed cluster centers
$c_j$ refer to the centers' positions at the end of the previous iteration.
The given equations for the bounds follow directly from the triangle inequality (Eq.~\ref{eq:tri}).

The calculation of these bounds entails an overhead of $\mathrm{O}(k^2)$ distance calculation per iteration.
However, this additional cost tends to be small compared to the number of distance calculations $\mathrm{O}(k\cdot|D|)$ that can potentially saved,
as  $k^2 \ll |D|$ in most practical applications.

>>>>

In this section, we first define the $k$-means clustering problem and the Lloyd algorithm as a straightforward, but optimizable solution to this problem in Section \ref{subsec: kmeans}.
We then continue with the introduction of means of acceleration trough lower and upper bounding in Section \ref{sub:acc}.

@@

In order to discuss the \kmeans, we first define the concept of a \emph{clustering algorithm} in a formal and implementation-agnostic manner:
\begin{definition}[clustering algorithm]
	Given a target number of clusters $k \in \mathbb{N}$, a clustering algorithm is a function $F$ that assigns each element from an input data set $D\subset \mathbb{R}^n$ a
	cluster index $i \in \{1, 2, \ldots, k\}$
	$$ F:D \to \{1, 2, \ldots, k\} \;.$$
\end{definition}
The intention is to assign similar objects the same index, thus clustering them together.
Notably, we restrict our analysis to clustering problems that use the $n$-dimensional Euclidean space as the domain of the input data set.
Likewise, we assume that the dissimilarity of the data points to be clustered is completely described by the Euclidean distance $d(x,y)= || x-y ||_2 $.
The rationale behind this assumption will be made clear shortly in Section \ref{sub:acc}.

The core of the \kmeans, then, is the assignment of clusters so that the variance in dissimilarity between all points in the same cluster should be reasonably small.
More formally:
\begin{definition}[\kmeans]
	Let $F$ be a clustering algorithm
	%$x \in D$ be an element from the input data set of a clustering algorithm $F$
	and $C_i$ be the set of all elements assigned to cluster $i$:
	$$ C_i = \{x \in D \mid F(x) = i\} \,.$$
	Furthermore, let $\epsilon$ be the weighted sum of inter-cluster variances
	$$ \epsilon = \sum_{i=1}^k |C_i| \operatorname{Var}[C_i] = \sum_{i=1}^k \sum_{x_j \in C_i}  d^2(x_j, c_i)\,, $$
	where the cluster's average element $c_i = \operatorname{E}[C_i]$ is called the center of the cluster $C_i$.

	Then, the map $F$ is called a solution to the \kmeans
	iff $\epsilon$ cannot be made smaller by changing the assignment of any one element to a different cluster.
\end{definition}
Note that with this definition, common $k$-means algorithms provide solutions to this \kmeans,
as we do not require that $F$ achieves the smallest possible $\epsilon$,
just a local minimum of $\epsilon$.
This is consistent with the colloquial notion that solutions to the \kmeans ``can be found'',
even when an optimal solution (an NP-hard problem) is not practically obtainable \cite{han2012data}.

The most prevalent algorithm employed to identify valid solutions to the \kmeans is the Lloyd algorithm \cite{DBLP:journals/tit/Lloyd82}, which is reproduced in Algorithm~\ref{alg:lloyd};
given its widespread use, we will only cover its most significant characteristics as they relate to our analysis.

@@

For each iteration of Lloyd's algorithm, the assignment step (line~\autoref{algstep:assign}) needs to identify the closest cluster center $c_j$ for each element $x_i$ from the input dataset $D$.
To do so, Lloyd's algorithm calculates the distances between each cluster and each element from $D$, requiring $k\cdot|D|$ total distance evaluations per iteration.
These frequent distance evaluations can make up a significant share of the algorithms computational costs, especially for high-dimensional datasets.

@@

As we have seen, Lloyd's algorithm requires numerous evaluations of the distance function.
However, the exact distances do not actually need to be calculated explicitly for every element, as it is sufficient to identify which cluster center is the closest to a given element.
This is often possible from geometric considerations alone, when lower and upper distance bounds are available.
Two properties of the Euclidean space are of particular use here:
The triangle inequality
\begin{align}
	\label{eq:tri}
	d(x,y) \leq d(x,z) + d(z,y)
\end{align}
and Ptolemy's inequality
\begin{align}
	\label{eq:pto}
	d(x, y)\cdot d(v, u) \leq d(x, v) \cdot d(y,v) + d(x, u) \cdot d(y, v)\;.
\end{align}
Both inequalities, known since ancient times, provide lower and upper bounds on the distances between two points given two (respectively five) other distances.

Numerous algorithms have been developed that exploit the former inequality to avoid explicit distance evaluations. However, to the best of our knowledge, the latter has not been employed for the purposes of accelerating $k$-means clustering yet.

To demonstrate the potential of the latter, we modify Elkan's algorithm \cite{DBLP:conf/icml/Elkan03} that already employs the triangle inequality, to also make use of Ptolemy's inequality.
We provide a short summary of Elkan's algorithm here and describe our extension in Section \ref{sec:contrib}.

Elkan's algorithm maintains two kinds of distance bounds in each iteration of the algorithm:
\begin{enumerate}[label=\roman*]
	\item $u(x) \geq d(x, c_i) \mid i = F(x)$,
	      upper bounds between data points and their currently assigned centers.
	\item $l(x, c_i) \leq d(x,c_i) \forall i \neq F(x)$,
	      lower bounds between points and all other centers.
\end{enumerate}
Intuitively, when the upper bound is larger than all lower bounds $u(x)\geq l(x,c_i) \forall i\neq F(x)$, the cluster assignment of a data point has not changed.
Explicit distances to other clusters are only required when this inequality does not hold, i.e. when the bounds are not tight enough or when the cluster assignment of the data point has indeed changed.
It is immediately evident that the quality of the bounds has a significant impact on the achievable improvements in execution speed:
The larger (smaller) the lower (upper) bound is, the more likely an explicit distance evaluation can be avoided.

While Elkan's algorithm also employs additional techniques to reduce the number of distance calculations, we want to focus on the calculation of the bounds here;
a complete description of the algorithm is given in Algorithm~\ref{alg:elkan}, as reproduced from \cite{DBLP:conf/icml/Elkan03}.

The lower and upper bounds are calculated at the end of each iteration of Elkan's algorithm and are given by
\begin{align}
	\label{eq:elkan_lower}
	l'(x_i, c_j) & = \max \{ l(x_i, c_j) - d(c'_j, c_j), 0 \} \\
	\label{eq:elkan_upper}
	u'(x_i)      & = u(x_i) + d(c'(x_i), c(x_i)) \,,
\end{align}
where the prime symbol ($c'_j, u', l'$) indicates quantities calculated in this iteration, while unprimed variables refer to quantities calculated in the preceding iteration.
The given equations for the bounds follow directly from the triangle inequality~\ref{eq:tri}.

The calculation of these bounds entails an overhead of $\mathrm{O}(k^2)$ distance calculations per iteration.
However, this additional cost tends to be small compared to the number of distance calculations $\mathrm{O}(k\cdot|D|)$ that can potentially saved,
as  $k \ll |D|$ in most practical applications.

### 2024-11-14 content/5_results.tex
overleaf version as submitted

In order to demonstrate the efficacy of our proposed algorithm, we undertake a comparative analysis with the performance of Elkan's algorithm, which it extends.
We measure the number of distance evaluations for multiple datasets and for different values of $k$, the number of clusters.
Our implementation is available through GitHub~\cite{Averitchev2024kMeansPtolemy}.

@@

\begin{table}[t]
	\centering
	\caption{Datasets used in the experiments}
	\label{tab:datasets}
	\begin{tabular}{llSS}
		\toprule
		\textbf{Group}                                    & \textbf{Dataset}          & \textbf{Cardinality} & \textbf{Dimensionality} \\
		\midrule
		\multirow{2}{*}{\textit{fixed structure}}         & Iris                      & 150                  & 4                       \\
		                                                  & Wine                      & 178                  & 13                      \\
		\cmidrule{2-4}
		\multirow{6}{*}{\textit{variable dimensionality}} & Gaussian Low Dimension    & 10000                & 10                      \\
		                                                  & Gaussian Medium Dimension & 10000                & 100                     \\
		                                                  & Gaussian High Dimension   & 10000                & 300                     \\
		                                                  & Random Low Dimension      & 5000                 & 10                      \\
		                                                  & Random Medium Dimension   & 5000                 & 100                     \\
		                                                  & Random High Dimension     & 5000                 & 300                     \\
		\bottomrule
	\end{tabular}
\end{table}

@@

Additionally, two intermediate variants were implemented to assess the separate effects of the lower and upper bounds:
\emph{Ptolemy\_upper} one uses Elkan's lower bound with Ptolemy's upper bound, and \emph{Ptolemy\_lower} uses Ptolemy's lower bound with Elkan's upper bound.

@@

In \autoref{fig:combined}, we report our results as distance calculations saved relative to Elkan's algorithm, showcasing any improvements introduced by the proposed extension.

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth]{fig/combined_plot.pdf}
	\caption{
		Relative number of saved distance evaluations, compared to Elkan's algorithm.
		Top and middle row: Performance increases on synthetic data generated from Gaussian (top) and uniformly random (middle) distributions.
		The dimensionality of the dataset increases left-to-right.\\
		Bottom row: Performance increases on the real-world datasets Iris (left) and Wine (right).
	}
	\label{fig:combined}
\end{figure}

The primary focus of this analysis is the full Ptolemaic algorithm, as it is the most robust when comparison to Elkan's algorithm.
In general, the full Ptolemaic extension shows substantial improvements in reducing the number of distance evaluations across most scenarios.

@@

\textbf{Dimensionality.} The full Ptolemaic algorithm performs well across varying dimensionality. In low-dimensional datasets (10 dimensions), it achieves impressive speedups, particularly at higher cluster counts.
As dimensionality increases, performance gains persist but become slightly less pronounced.
Nonetheless, even in higher-dimensional datasets, the full Ptolemaic algorithm remains effective, consistently reducing distance evaluations.
However, the Ptolemaic lower bound variant often struggles in these higher-dimensional cases, further emphasizing the robustness of the full Ptolemaic approach.

@@

\textbf{Limitations.}
The presented experiments are subject to three primary constraints:
(i) While synthetic random datasets provide valuable best-case scenarios, more extensive real-world data with appropriate grouping structures is necessary for comprehensive evaluation.
(ii) The significant reduction in distance calculations may not directly translate to improved execution times, as Ptolemy's bounds are computationally more expensive than the triangle inequality.
(iii) Most critically, the limited sample size and fixed initial conditions, particularly the lack of variation in cluster initializations, severely restrict the generalizability of individual measurements.\\
Despite these limitations, we contend that the observed overall trend remains compelling and warrants further investigation into the application of Ptolemy's inequality for clustering algorithms.

>>>>

In order to demonstrate the efficacy of our proposed algorithm, we undertake a comparative analysis with the performance of Elkan's algorithm, which it extends.
We measure the number of distance evaluations for different benchmark datasets and for different values of $k$, the number of clusters.
Our implementation is available through GitHub~\cite{Averitchev2024kMeansPtolemy}.

@@

\begin{table}[t]
	\centering
	\caption{Datasets used in the experiments.}
	\label{tab:datasets}
	\begin{tabular}{llSS}
		\toprule
		\textbf{Group}                                    & \textbf{Dataset}          & \textbf{Cardinality} & \textbf{Dimensionality} \\
		\midrule
		\multirow{2}{*}{\textit{fixed structure}}         & Iris                      & 150                  & 4                       \\
		                                                  & Wine                      & 178                  & 13                      \\
		\cmidrule{2-4}
		\multirow{6}{*}{\textit{variable dimensionality}} & Gaussian Low Dimension    & 10000                & 10                      \\
		                                                  & Gaussian Medium Dimension & 10000                & 100                     \\
		                                                  & Gaussian High Dimension   & 10000                & 300                     \\
		                                                  & Random Low Dimension      & 5000                 & 10                      \\
		                                                  & Random Medium Dimension   & 5000                 & 100                     \\
		                                                  & Random High Dimension     & 5000                 & 300                     \\
		\bottomrule
	\end{tabular}
\end{table}

@@

Additionally, two intermediate variants were implemented to assess the separate effects of the lower and upper bounds:
The \emph{Ptolemy Upper Bound} method uses Elkan's lower bound and Ptolemy's upper bound, while \emph{Ptolemy Lower Bound} uses Ptolemy's lower bound and Elkan's upper bound.

@@

In \autoref{fig:combined},
we report our results in the form of speedups,
i.e. as the number of distances calculated by our extension,
divided by the number of distances calculated by Elkan's algorithm.

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth]{fig/combined_plot.pdf}
	\caption{
		Relative improvement in the number distance evaluations, compared to Elkan's algorithm.
        A speedup of $>1$ indicates an improvement over Elkan's original algorithm.
		\textbf{Top and middle row:} Performance increases on synthetic data generated from Gaussian (top) and uniformly random (middle) distributions.
		The dimensionality of the dataset increases left-to-right.
		\textbf{Bottom row:} Performance increases on the real-world datasets Iris (left) and Wine (right).
	}
	\label{fig:combined}
\end{figure}

The primary focus of this analysis is the full Ptolemaic algorithm as compared to Elkan's algorithm, since it is the most robust one of the three variants. 
In general, the full Ptolemaic extension shows substantial improvements in reducing the number of distance evaluations across most scenarios.

@@

\textbf{Dimensionality.} The full Ptolemaic algorithm performs well across varying dataset dimensionality. In low-dimensional datasets (10 dimensions), it achieves impressive speedups, particularly at higher cluster counts.
As dimensionality increases, performance gains persist but become slightly less pronounced.
Nonetheless, even in higher-dimensional datasets, the full Ptolemaic algorithm remains effective, consistently reducing distance evaluations.
However, the Ptolemaic lower bound variant often struggles in these higher-dimensional cases, further emphasizing the robustness of the full Ptolemaic approach.

@@

\textbf{Limitations.}
The presented experiments are subject to three primary constraints:
(i) While synthetic random datasets provide valuable best-case scenarios, more extensive real-world data with appropriate grouping structures is necessary for comprehensive evaluation.
(ii) The significant reduction in distance calculations may not directly translate to improved execution times, as Ptolemy's bounds are computationally more expensive than the triangle inequality.
(iii) Most critically, the limited sample size and fixed initial conditions, particularly the lack of variation in cluster initializations, severely restrict the generalizability of individual measurements.

Despite these limitations, we conclude that the observed overall trend remains compelling and warrants further investigation into the application of Ptolemy's inequality for clustering algorithms.

### 2024-11-14 main.tex
overleaf version as submitted

\maketitle
\begin{abstract}
	Clustering is a fundamental data analytics operation in the field of unsupervised learning. Given a database of unknown structure, clustering aims to discover the inherent structure of the data objects according to similarity, such that similar data objects are grouped together, while dissimilar ones are separated in different groups or clusters.
	In this study, we focus on partitioning methods, specifically the \emph{k-means clustering} approach, which minimizes the intra-cluster variance.
	As the standard algorithmic approach to k-means clustering, namely the Lloyd algorithm, is neither efficient nor scalable, various adaptations and modifications have been developed, resulting in the family of fast k-means clustering algorithms. In this short paper, we extend the clustering algorithm proposed by Elkan with Ptolemy's inequality to prune superfluous distance calculations. It is not our intention with this paper to compete with the state-of-the-art algorithmic solutions for the k-means clustering problem; rather, we seek to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, particularly in terms of reducing computations associated with distance evaluations.

@@

%intro
Clustering is a fundamental data-analytical operation in the domain of data science.
The basic task is to divide a set of data objects into different groups or clusters, such that objects within the same cluster have sufficient similarities;
meanwhile objects in different clusters should have significant differences.
The collection of clusters reflects the inherent structure of the database and is denoted as clustering.

Clustering can be found in various scientific or economic application domains \cite{ezugwu2022comprehensive,oyewole2023data, gan2020data}, ranging from archaeology \cite{troiano2024comparative} and finance \cite{cai2016clustering} to industry \cite{lee2021technological} and zoology \cite{shen2021multivariate}, to name just a few. Especially in the field of unsupervised learning, clustering is utilized to extract structures from unlabeled data \cite{chander2023data}.

Alongside the diverse applications of clustering, unique domain-specific requirements and challenges arise, each of which can be addressed by leveraging different families of clustering approaches \cite{xu2015comprehensive,han2012data}. These approaches, ranging from partitioning and hierarchical methods to density-based, grid-based and graph-based methods, are designed to accommodate varying data characteristics and application-related requirements to facilitate efficient cluster analyses across diverse domains.

@@

With the proliferation of complex data sources and the growing demand for more efficient clustering techniques, numerous \emph{fast k-means algorithms} have been introduced \cite{DBLP:conf/icml/Elkan03,DBLP:conf/sdm/Hamerly10,drake2012accelerated,hamerly2015accelerating,DBLP:conf/icml/NewlingF16,DBLP:conf/icml/DingZSMM15,DBLP:conf/sisap/SchubertLF21,DBLP:conf/sisap/LangS23}. These methods are designed to achieve exact clustering results with significantly reduced computational cost. Moreover, most of them do not require any form of precomputation, making them directly applicable across a wide range of scenarios. The major objective of fast k-means algorithms is to reduce the number of distance evaluations required when assigning data objects to cluster centers and to safely exclude cluster centers from the assignment process. For this purpose, distances are approximated via lower and upper bounds.

The Elkan algorithm \cite{DBLP:conf/icml/Elkan03} is a prominent representative of these algorithms and particularly suited for high-dimensional scenarios.
It maintains one upper bound and multiple lower bounds for each data object in order to apply different pruning criteria for each combination of data object and cluster center.
While the lower and upper bounds of Elkan's algorithm are directly derived from the triangle inequality, we propose to utilitze Ptolemy's inequality instead in order to increase the pruning performance.
With this extension, we do not aim to compete with the state of the art in k-means clustering;
rather we intend to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, specifically in terms of reducing computations associated with distance evaluations. We believe that our research findings offer promising directions for future research and new research impulses for data scientists and practitioners alike.

This paper is structured as follows:
Section~\ref{Preliminaries} outlines the k-means problem and Elkan's approach of acceleration the algorithmic computation.
In Section~\ref{sec:contrib}, we proposes how to apply Ptolemy's inequality to the aforementioned algorithm as our main contribution.
The results of our preliminary performance evaluation are detailed in Section~\ref{sec: results}, while Section~\ref{sec: conclusions} concludes this paper with a short outlook on future work.

@@

Moreover, the use of the Ptolemy's inequality opens up new avenues for future work.
Besides conducting more extensive experiments to verify our findings,
it would be valuable to explore adaptations of this technique for other k-means algorithms and optimizations.
In particular for algorithms which rely on the triangle inequality to prune distance calculations, Ptolemy's inequality can potentially be leveraged to achieve an enhancement in performance, making these clustering algorithms more practical for broader range of applications.

>>>>

\maketitle
\begin{abstract}
	Clustering is a fundamental data analytics operation in the field of unsupervised learning. Given a database of unknown structure, clustering aims to discover the inherent structure of the data objects according to similarity, such that similar data objects are grouped together, while dissimilar ones are separated in different groups or clusters.
	In this study, we focus on partitioning methods, specifically the \emph{k-means clustering} approach, which minimizes the intra-cluster variance.
	As the standard algorithmic approach to k-means clustering, the Lloyd algorithm, is neither efficient nor scalable, various adaptations and modifications have been developed, resulting in the family of fast k-means clustering algorithms.\\
    In this short paper, we extend the clustering algorithm proposed by Elkan with Ptolemy's inequality to prune superfluous distance calculations. It is not our intention to compete with the state-of-the-art algorithmic solutions for the k-means clustering problem; rather, we seek to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, particularly in terms of reducing computations associated with distance evaluations.

@@

%intro
Clustering is a fundamental data-analytics operation in the domain of data science.
The basic task is to divide a set of data objects into different groups or clusters, such that objects within the same cluster have sufficient similarities;
meanwhile objects in different clusters should have significant differences.
The collection of clusters reflects the inherent structure of the underlying data-generating process and is denoted as a clustering.

The need for clustering arises in various scientific or economic application domains \cite{ezugwu2022comprehensive,oyewole2023data, gan2020data}, ranging from archaeology \cite{troiano2024comparative} and finance \cite{cai2016clustering} to industry \cite{lee2021technological} and zoology \cite{shen2021multivariate}, to name just a few. Especially in the field of unsupervised learning, clustering is utilized to extract structures from unlabeled data \cite{chander2023data}.

Alongside the diverse applications of clustering, unique domain-specific requirements and challenges arise, which are addressed by leveraging different families of clustering approaches \cite{xu2015comprehensive,han2012data}. These approaches, ranging from partitioning and hierarchical methods to density-based, grid-based and graph-based methods, are designed to accommodate varying data characteristics and domain-related requirements to facilitate efficient cluster analyses across diverse applications.

@@

With the proliferation of complex data sources and the growing demand for more efficient clustering techniques, numerous \emph{fast k-means algorithms} have been introduced \cite{DBLP:conf/icml/Elkan03,DBLP:conf/sdm/Hamerly10,drake2012accelerated,hamerly2015accelerating,DBLP:conf/icml/NewlingF16,DBLP:conf/icml/DingZSMM15,DBLP:conf/sisap/SchubertLF21,DBLP:conf/sisap/LangS23}. These methods are designed to achieve good clustering results with significantly reduced computational cost. Moreover, most of them do not require any form of precomputation, making them directly applicable across a wide range of scenarios. The major objective of fast k-means algorithms is to reduce the number of distance evaluations required when assigning data objects to cluster centers and to safely prune cluster centers from the assignment process. For this purpose, distances are approximated via lower and upper bounds.

The Elkan algorithm \cite{DBLP:conf/icml/Elkan03} is a prominent representative of these algorithms and particularly suited for high-dimensional scenarios.
It maintains one upper bound and multiple lower bounds for each data object in order to apply different pruning criteria for each combination of data object and cluster center.
While the lower and upper bounds of Elkan's algorithm are directly derived from the triangle inequality, we propose to utilitze Ptolemy's inequality instead in order to increase the pruning performance.
With this extension, we do not aim to compete with the state of the art in k-means clustering;
rather, we intend to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, specifically in terms of reducing computations associated with distance evaluations. We believe that our research findings offer a promising direction for future research and are beneficial for data scientists researchers and practitioners alike.

This short paper is structured as follows:
Section~\ref{Preliminaries} outlines the k-means problem and Elkan's approach of accelerating the algorithmic computation.
In Section~\ref{sec:contrib}, we show how Ptolemy's inequality can be applied to the aforementioned algorithm. %as our main contribution.
The results of our preliminary performance evaluation are detailed in Section~\ref{sec: results}, while Section~\ref{sec: conclusions} concludes this paper with a short outlook on future work.

@@

Moreover, the use of the Ptolemy's inequality opens up new avenues for future work.
Besides conducting more extensive experiments to verify our findings,
it would be valuable to explore adaptations of this technique for other k-means algorithms and optimizations.
In particular for algorithms which rely on the triangle inequality to prune distance calculations, Ptolemy's inequality can potentially be leveraged to achieve an enhancement in performance, making these clustering algorithms more practical for a broader range of applications.
